Comparison of Several Stochastic and Deterministic Derivative-free Global Optimization Algorithms

### 2019-07-15 2019/MOTOR_Sovrasov/slides/slides.tex
MOTOR: add the final version of slides

\begin{frame}
  \frametitle{Methods}
  In this work considered the following solvers available in open-source:
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item DIRECT, DIRECT$l$;
      \item AGS, AGS$l$.
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Multi Level Single Linkage;
      \item StoGO;
      \item Differential Evolution;
      \item Controlled Random Search;
      \item Dual Simulated Annealing;
      \item Ant Colony Optimization.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Univariate Algorithm of Global Search}
  Optimization method generates search sequence \(\{x_k\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose \(p\) intervals \((x_{t_j-1}, x_{t_j})\) with the greatest characteristics and
    compute objective \(\varphi(y(x^{k+j}))\) in points chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1+j}=d(t)\in (x_{t_j-1}, x_{t_j}),\:j=\overline{1,p}
    \end{displaymath}
    \item[Step 4.] If \(x_{t_j}-x_{t_j-1}<\varepsilon\) for one of \(j=\overline{1,p}\), stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

@@

\begin{frame}
  \frametitle{Hyperparameters tuning in AGS}
  AGS selects the best intervals according to the following characteristic:
  \begin{displaymath}
    R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{(rH)^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{rH},1<i<k+1
  \end{displaymath}
  where $L$ is current estimation on Holder constant, $r$ is reliability coefficient.
  AGS is very sensitive to choice of $r$:
  \begin{itemize}
    \item low value can lead to fast convergence to a local minima
    \item high value leads to very slow convergence
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Test problems}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      Generator GKLS was employed to construct 8 sets of 100 test problems:
      \begin{displaymath}
        \begin{matrix}
          f(x)=
          \left\{
          \begin{matrix}
          C_i(x), x \in S_i, i\in 2,\dots ,m \\
          \Vert x-T \Vert^2 + t, x\not\in S_2,\dots,S_m
          \end{matrix} \right.
        \end{matrix}
      \end{displaymath}

@@

\begin{frame}
  \frametitle{Experimental setup}
  \begin{table}
  \begin{center}
  \caption{Class-specific parameters of the optimization algorithms}
    \begin{tabular}{|l|{c}|{c}|{c}|}
      \hline
      & AGS & CRS & DE\\
    \hline
    \(F_{GR}\) & \(r=3\) & popsize=150 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Simple & \(r=4.6\) & popsize=200 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Hard & \(r=6.5\) & popsize=400 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 3d Simple & \(r=3.7\) & popsize=1000 & mutation=(1.1,1.9), popsize=70 \\
    \hline
    GKLS 3d Hard & \(r=4.4\) & popsize=2000 & mutation=(1.1,1.9), popsize=80 \\
    \hline
    GKLS 4d Simple & \(r=4.7\) & popsize=8000 & mutation=(1.1,1.9), popsize=90 \\
    \hline
    GKLS 4d Hard & \(r=4.9\) & popsize=16000 & mutation=(1.1,1.9), popsize=100 \\
    \hline
    GKLS 5d Simple & \(r=4\) & popsize=25000 & mutation=(1.1,1.9), popsize=120 \\
    \hline
    GKLS 5d Hard & \(r=4\) & popsize=30000 & mutation=(1.1,1.9), popsize=140 \\
    \hline
  \end{tabular}
    \label{tab:params}
  \end{center}
  \end{table}
  \begin{itemize}
    \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
    \item in the SDA method, the parameter \(visit=2.72\).
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Robustness of the AGS-AR to hyperparemeters}
  \begin{figure}[ht]
    \centering
    \includegraphics[width=.6\textwidth]{ar_stab.pdf}
    \caption{Operating characteristics of AGS and AGS-AR with different hyperparameters
    when solving problems from the GKLS 5d Simple classe.}
    \label{fig:stability}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item The proposed modification of the stock AGS, AGS-AR allows to pay less attention to initial hyperparameter tuning and
      performs on-par with properly tuned AGS;
      \item AGS-AR method has demonstrated the convergence
      speed and reliability at the level of DIRECT and exceeds many other algorithms, the open-source
      implementations of which are available in web;
      \item the stochastic optimization methods inferior to the deterministic ones in the convergence
    speed and in reliability. It is manifested especially strongly on more complex multiextremal
    problems.
    \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Methods}
  In this work considered the following solvers available in open-source:
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}:
    \begin{itemize}
      \item DIRECT, DIRECT$l$ (Gablonsky J. M. et al, 2001);
      \item AGS (Strongin R.G., 1978).
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}:
    \begin{itemize}
      \item Multi Level Single Linkage (Alexander H. G. et al, 1987);
      \item StoGO (Madsen, Kaj at al, 1998);
      \item Differential Evolution (Storn, Rainer et al, 1997);
      \item Controlled Random Search (Price, W. L., 1983);
      \item Dual Simulated Annealing (Y Xiang et al, 1997);
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Univariate Algorithm of Global Search}
  Optimization method generates search sequence \(\{x_k:x_k\in[a,b]\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose the interval \((x_{t-1}, x_{t})\) with the greatest characteristic and
    compute objective \(\varphi(y(x^{k+1}))\) in the point chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1}=d(t)\in (x_{t-1}, x_{t})
    \end{displaymath}
    \item[Step 4.] If \(x_{t}-x_{t-1}<\varepsilon\) stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

@@

\begin{frame}
  \frametitle{Hyperparameters tuning in AGS}
  AGS selects the best intervals according to the following characteristic:
  \begin{displaymath}
    R(i)=\Delta_i+\dfrac{(z_i-z_{i-1})^2}{(rH)^2\Delta_i}-2\dfrac{z_i+z_{i-1}}{rH},1<i<k+1
  \end{displaymath}
  where $L$ is the current estimation of the Holder constant, $r$ is reliability coefficient.
  AGS is very sensitive to choice of $r$:
  \begin{itemize}
    \item low value can lead to fast convergence to a local minima
    \item high value leads to very slow convergence
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Test problems}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      Generator GKLS (Gaviano, M. et al, 2003) was employed to construct 8 sets of 100 test problems:
      \begin{displaymath}
        \begin{matrix}
          f(x)=
          \left\{
          \begin{matrix}
          C_i(x), x \in S_i, i\in 2,\dots ,m \\
          \Vert x-T \Vert^2 + t, x\not\in S_2,\dots,S_m
          \end{matrix} \right.
        \end{matrix}
      \end{displaymath}

@@

\begin{frame}
  \frametitle{Experimental setup}
  \begin{table}
  \begin{center}
  \caption{Class-specific parameters of the optimization algorithms}
    \begin{tabular}{|l|{c}|{c}|{c}|}
      \hline
      & AGS & CRS & DE\\
    \hline
    \(F_{GR}\) & \(r=3\) & popsize=150 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Simple & \(r=4.6\) & popsize=200 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 2d Hard & \(r=6.5\) & popsize=400 & mutation=(1.1,1.9), popsize=60 \\
    \hline
    GKLS 3d Simple & \(r=3.7\) & popsize=1000 & mutation=(1.1,1.9), popsize=70 \\
    \hline
    GKLS 3d Hard & \(r=4.4\) & popsize=2000 & mutation=(1.1,1.9), popsize=80 \\
    \hline
    GKLS 4d Simple & \(r=4.7\) & popsize=8000 & mutation=(1.1,1.9), popsize=90 \\
    \hline
    GKLS 4d Hard & \(r=4.9\) & popsize=16000 & mutation=(1.1,1.9), popsize=100 \\
    \hline
    GKLS 5d Simple & \(r=4\) & popsize=25000 & mutation=(1.1,1.9), popsize=120 \\
    \hline
    GKLS 5d Hard & \(r=4\) & popsize=30000 & mutation=(1.1,1.9), popsize=140 \\
    \hline
  \end{tabular}
    \label{tab:params}
  \end{center}
  \end{table}
  \begin{itemize}
    \item in the DIRECT and DIRECT\(l\) methods, parameter \(\epsilon=10^{-4}\);
    \item in the SDA method, the parameter \(visit=2.72\).
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Robustness of the AGS-AR to hyperparemeters}
  \begin{figure}[ht]
    \centering
    \includegraphics[width=.6\textwidth]{ar_stab.pdf}
    \caption{Operating characteristics of AGS and AGS-AR with different hyperparameters
    when solving problems from the GKLS 5d Simple class.}
    \label{fig:stability}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Conclusions}
    %Already done:
    \begin{itemize}
      \item The proposed modification of the stock AGS, AGS-AR allows to pay less attention to initial hyperparameter tuning and
      performs on-par with properly tuned AGS;
      \item AGS-AR method has demonstrated convergence
      speed and reliability at the level of DIRECT and exceeds many other algorithms, open-source
      implementations of which are available in web;
      \item considered stochastic optimization methods inferior to the deterministic ones in the convergence
    speed and in reliability. It is manifested especially strongly on more complex multiextremal
    problems.
    \end{itemize}
\end{frame}
